---
title: "A Model Statistics Course"
author: "Daniel Kaplan"
date: "2023-10-18"
categories: [presentation, overview]
format:
  html:
    toc: true
---

> *These are presentation notes for the October 2023 StatChat meeting. For more than 15 years, statistical educators in the Twin Cities region of Minnesota have been gathering a half-dozen times a year at StatChat to share comradeship and teaching insights. Among the schools regularly represented are the University of Minnesota, Macalester College, St. Olaf College, Hamline University, Augsburg University, Carleton College, St. Cloud State University, and Minnesota State University Mankato.*

**Abstract**: "Mere Renovation is Too Little Too Late: We Need to Rethink Our Undergraduate Curriculum from the Ground Up" is the title [2015 paper](https://nhorton.people.amherst.edu/mererenovation/) by George Cobb. Honoring George's challenge, I have been rethinking and re-designing the introductory statistics course, replacing traditional foundations using modern materials and reconfiguring the living and working spaces to suit today's applied statistical needs and projects. In the spirit of a "model house" used to demonstrate housing innovations, I'll take you on a tour of my "model course," whose materials are available free, open, and online. Among the features you'll see: an accessible handling of causal reasoning, a unification of the course structure around modeling, a streamlined yet professional-quality computational platform, and an honest presentation of Null Hypothesis Testing that connects it to more current forms of statistical inference.

```{r include=FALSE}
library(LST)
```


## Motivation

The "consensus" Stat 101 is 50 years out of date:

- fails to engage issues of causation, covariation, and adjustment
- too much emphasis on p-values 
- entirely ignores Bayes
- no substantial coverage of risk, risk factors, ... 
- uses a confusing over-variety of graphic modes (many of which are out-of-date)
- doesn't make contact with data science, machine learning, and AI/GPT

My objective: Demonstrate the extent to which it's possible to overcome these deficiencies with a complete, practicable, no-prerequisite course.

I'm happy to discuss the above points anytime, but that's not the point of this talk.

::: {.callout-note}
## Jeff Witmer's approach

Jeff proposes 15 changes*, dividing into effort categories:

a. Changes You Could Make with Little Effort or Planning. (e.g. "significant" -> discernible)
b. Changes to a Course That You Could Implement after Investing a Day or so of Planning. (e.g. emphasize power, not $\alpha$)
c. Changes to a Course That Would Require Quite a Bit of Planning but That Are Worth Considering Nonetheless (e.g. emphasize effect size)

Jeff Witmer (2023) "What Should We Do Differently in STAT 101?" *Journal of Statistics and Data Science Education* [link](https://www.tandfonline.com/doi/full/10.1080/26939169.2023.2205905)

*Almost all of which are engaged in *Lessons in Statistical Thinking*
:::

## Style

a. Demonstrate and describe statistical phenomena by causal simulation
    - Examples:
        - sampling variation
        - confounders, covariates, colliders, adjustment
    - Stat theory from simulation/wrangling rather than probability/algebra
#. Informal inference from the very beginning, *gradually* formalizing it over the semester
#. Single, standard format for graphics


## Course overview

- Part 1: Handling data
    - Data frames
    - Graphics (data and models)
    - Wrangling

- Part 2: Describing relationships
    - Regression (incl. categorical and multiple explanatory variables)
    - Adjustment

- Part 3: Randomness and the unexplained
    - Signal and noise
    - Simulation and DAGs
    - Probability models (optional)
    - Likelihood (optional, prep. for Part 5)
    - Measuring and accumulating risk

- Part 4: Understanding modeling
    - Sampling variation and confidence intervals/bands 
    - Causality/Confounding/Adjustment
    - Experiment

- Part 5: Hypothetical thinking
    - Basic Bayes: competing two hypotheses
    - Null hypothesis testing (What you get when there is no prior and no second hypothesis.)

## How can we fit more in an already crowded course?

Streamline!

1. Reduce ~~drag~~ cognitive load. 
    a. Repeated use a small number of standard forms
        - one basic graphical pattern: annotated point plot
        - one basic computational pattern: verb `|>` verb `|>` ...
    b. Avoid nomenclature conflicts with everyday words, e.g.
        - "table" -> data frame
        - "case" -> specimen
        - "assignment" -> storage
2. Unify t, p into regression modeling, conf. interval/bands in both graphics and models
3. Keep number of types of objects small: data frame, model, graphic, simulation
    a. Very small computational footprint, a dozen stat/graphics/wrangling functions.
4. Remove square roots whenever that's easy
    - focus on variance rather than standard deviation, R^2 rather than r


## Lesson 1. Data frames

Data is *always* in data frames.

- Columns: Variables

- Rows: "Specimens" / Unit of observation

Computing concepts:

- name of data frame
- pipe
- function()

Usually start with a named data frame, piping it to a function.

```{r}
Nats |> names()
```

## Lesson 2. Graphics

Both the horizontal and vertical axes are mapped to variables.

Just one command: `pointplot()` produces point plot with automatic jittering as needed.

Tilde expression specifies which variable is mapped to y and x (and, optionally, color and faceting).

```{r}
#| layout-ncol: 2
Galton |> pointplot(height ~ sex)
Galton |> pointplot(height ~ mother + sex + sex)
```

## Lesson 3. Empirical distributions

```{r}
Galton |> pointplot(height ~ sex, annot = "violin")
```

## Lesson 4. Models as graphical annotation

```{r}
#| layout-ncol: 2
Galton |> sample_n(size=100) |> 
  pointplot(height ~ sex, annot = "model", alpha = 0.1, model_alpha=0.75)
Galton |> pointplot(height ~ mother + sex + sex, annot = "model", alpha = 0.1, model_alpha=0.75)
```

## Lesson 5: Wrangling

[Perhaps use two class days]

Five basic operations: `mutate()`, `filter()`, `summarize()`, `select()`, `arrange()`


```{r}
Nats
```

```{r}
Nats |> filter(year == 2020)
Nats |> summarize(totalpop = sum(pop), .by=year)
```

## Lesson 6: Computing recap

[Perhaps merged into a two-day wrangling unit with Lesson 5]

Pipes, functions, parentheses, arguments, ...

## Lesson 7: Databases

[Entirely optional]

- Joins 
- Why we put related data into tables with different units of observation.



## Objects and operations

a. Data frame
#. Data graphics (as distinct from "infographics")
#. Statistical model
#. Simulation

Operations for all students

1. Data wrangling (simplified)
2. Annotated point plot of variables from a data frame
3. Model training
4. Model summarization

Operations used in demonstrations (and suited to some students)

5. Simulation (in demonstrations)
6. Iteration and accumulation (in demonstrations)

Computations on variables are always **inside** the arguments of a function taking a data frame as an input.

Tilde expressions for models and graphics.

## Resources

1. Textbook: [**Statistical Inference via Data Science**](https://moderndive.com) by Chester Ismay and Albert Y. Kim
2. Textbook: [**Lessons in Statistical Thinking**](https://dtkaplan.github.io/Lessons-in-statistical-thinking) by Danny Kaplan
    - [Associated R package: `{LST}` ](https://dtkaplan.github.io/LST/)
3. [USAFA Math 300Z](https://dtkaplan.github.io/Math-300Z/) course outline, instructor notes, etc.
4. Jeff Witmer (2023) "What Should We Do Differently in STAT 101?" *Journal of Statistics and Data Science Education* [link](https://www.tandfonline.com/doi/full/10.1080/26939169.2023.2205905)



## Basement

- [ ] one
- [ ] two


Definition  
: Definition is a statement that describes the meaning of the listed word in the heading.   

Hypothesis
: A hypothesis is a statement that might or might not be true.

