---
title: "A bad graph for medical screening"
author: "Daniel Kaplan"
date: "2023-04-30"
categories: [prevalence, sensitivity, specificity, prior, posterior]
---

```{r include=FALSE}
library(math300)
source("../../../_startup.R")

```

In Lesson 35, in the context of medical screening tests, we presented diagrams like this one.

```{r echo=FALSE, warning=FALSE}
library(ggmosaic)
med_test_graph(prevalence=0.15, sensitivity=0.8, specificity=0.7, n=10000)
```

This diagram is based on only three basic numbers---sensitivity, specificity, and prevalence. Exactly the same information could be presented in a 2x2 table:

Test result | Sick patients | Healthy patients
--------|--------------|------------
$\mathbb P$ | 12% (true positives) | 26% (false positives)
$\mathbb N$ | 3% (false negatives) | 60% (true negatives)

The four numbers necessarily add up to 100%, so one of the numbers is redundant. To generate the table we only need the three basic numbers:

- prevalence: 15%, that is, true positives + false negatives 
- sensitivity: 12%/(12%+3%) = 80%, that is, true positives divided by prevalence
- specificity: 70.6%, that is, true negatives/(1-prevalence). Filling in the numbers 60%/(1-15%) = 70.6%. 



Usually in statistical graphics, we place the scales on the horizontal and vertical axes, which is not the case with the above diagram. Sticking with the scales-on-axes convention, here is a streamlined graph:

```{r echo=FALSE}
hypothesis_compete_graph(prior=.15, like1=0.8, like2=0.3)
```

We've generalized the notation a bit and emphasized (1-specificity) rather than the specificity itself. 

1. Prior(Alternative) = width of "Alternative" box.
2. Likelihood for Alternative hypothesis, that is, p($\mathbb P$ | Alternative) (corresponds to sensitivity)
3. Likelihood for Null hypothesis, p($\mathbb P$ | Null) (corresponds to 1-specificity.)

The area of each box is, as expected, the width times the height. The two areas printed on the graph are the ingredients for the calculation of the posterior:

posterior(Alternative | $\mathbb P$) = 0.12/(0.12 + 0.255) = 32%
